A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-08-09 paper-hierarchical.tex
revised version

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF \cite{ID16A} in two scenarios. This beamline was built during the Upgrade Phase~I at ESRF (2009-2015). It uses one or several undulators as the source. During the period 2011-2018, it operated exploiting the ESRF-1 storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000~pm. In the second phase of the ESRF Upgrade, the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the X-ray sources will be much more coherent, especially in the horizontal direction.

ID16A is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. The beamline can operate at two fixed photon energies: 17050~eV and 33600~eV. The straight section in the magnet lattice houses four undulators, two 18~mm and another two of 22.4~mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

\begin{table}\label{tab:Undulators}
\centering
\caption{Parameters of the undulators available at the ID16A beamline: period ($\lambda_u$), length ($L$), number of identical undulators ($N$) and deflection parameter ($K$) values for the three main photon energies in use. The valuse are shown for the two ESRF magnetic lattices: the old ESRF-1, with $E_e = 6.04$~GeV), and the new ESRF-EBS, with $E_e = 6.00$~GeV. $n$ is the emission harmonic in use for each particular photon energy.}

\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
name  & $\lambda_u$ [mm] & $L$ [m] & N & $K$ (11200~eV) & $K$ (17050~eV) & $K$ (33600~eV) \\ \\
%\input{tableUndulator.txt}
 ESRF-1 U18.3  & 18.30  & 1.40  & 2 & 1.175(n=1) & 0.470(n=1) & 1.175(n=3)\\ 
 ESRF-1 U22.4  & 22.40  & 1.40  & 2 & 0.873(n=1) & 1.855(n=3) & 0.873(n=3)\\ \\
 EBS U18.3  & 18.30  & 1.40  & 2 & 1.156(n=1) & 0.437(n=1) & 1.156(n=3)\\ 
 EBS U22.4  & 22.40  & 1.40  & 2 & 0.852(n=1) & 1.836(n=3) & 0.852(n=3)\\ 
\end{tabular}%
}
\end{table}

@@

\begin{figure}\label{fig:ID16A}
    \centering
    \includegraphics[width=0.9\textwidth,clip=true,trim=50 85 50 85]{GRAPHICS/ID16A.pdf}
    \caption{Schematic of the ID16A beamline showing the horizontal (top) and vertical (bottom) planes and the position of the main optical elements. In the sketch ML stands for the focusing multilayer monochromator; VSS for virtual source slit; and KB for the Kirkpatrick-Baez set of mirrors. The sketch is drawn out of scale.}
\end{figure}

@@

\begin{table}\label{tab:eBeam}
    \centering
    \caption{ESRF-1 and EBS main electron beam parameters. Electron beam parameters for the high-$\beta$ (labelled ESRF) and the EBS straight sections. Values taken at the symmetry point of the straight section, where the insertion devices are placed. Values are taken from \cite{orangebook}, and updated for EBS with
    the lattice labelled S28D. 
    %For EBS, two (very close) design values are used, from Ref.~\cite{ESRF2014} labelled EBS-S28A, and those coming from further lattice refinement, labelled EBS-S28D. 
    %\todo{Remove ESRF-AT and EBS-S28A} 
    }
        \resizebox{\textwidth}{!}{%
        \begin{tabular}{ccccc}
         Lattice & $\sigma_e$ horizontal {[}$\mu$m{]} &  $\sigma_e'$ horizontal {[}$\mu$rad{]} & $\sigma_e$ vertical {[}$\mu$m{]} & $\sigma_e'$ vertical {[}$\mu$rad{]} \\ \\
         %\input{tableElectron.txt}
         ESRF-1  & 387.80  & 3.50  & 10.30 & 1.20 \\ 
         %ESRF-AT  & 414.97  & 3.43  & 10.31 & 1.16 \\ 
         %EBS-S28A  & 27.20  & 3.40  & 5.20 & 1.40 \\ 
         EBS     & 30.18  & 3.64  & 4.37 & 1.37 \\ 
    \end{tabular}
    }
\end{table}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$~eV: a value close to the usual operation value shown in Table \ref{tab:Undulators}. Calculations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF-1 where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.
This section is divided into three subsections, that present three main levels of simulations: simple calculations that can be made by hand, ray-tracing simulations and wave optics simulations: sections  \ref{level0}, \ref{level1} section \ref{level2} respectively.

\subsection{Analytical calculations}
\label{level0}

@@

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \frac{1}{2  \pi}\sqrt{2 \lambda L}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source has sizes $\Sigma$ and divergences $\Sigma'$ given by:
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

The beam sizes at different positions in the beamline are estimated in different ways depending upon whether the beam is focused or not. If the beam is focused, its dimensions, to a first approximation, are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam sizes in a position out of the focus the beam divergence is used. In an ideal {\'{e}}tendue-conserving optical system, where the Smith-Helmholz invariant applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

@@

%Effects of slope errors and diffraction limits can be simply estimated by hand calculations as well...
%The smearing of the focal size because of diffractive effects by the mirror clipping will also be discussed.

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$~eV for both the high-$\beta$ (ESRF-1) and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50~$\mu$m.
    %(21.3 if approximating by the Gaussian $\sigma$).
    }
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF-1 (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    FWHM$_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    FWHM$_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    FWHM'$_h$ [$\mu$rad]:   & 26.96         & 65.22        & 189062   & 15.60   & 37.74   & 109404       \\
    FWHM'$_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons in the sample. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most advanced and well-maintained codes used in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated using SRW over an aperture of $1 \times 1$~mm$^2$ placed at $d=27.1$~m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use.
%($\max(\Sigma') \times d = 0.73$~mm), with $\Sigma'$ values from in Table~\ref{tab:HandCalculations}. 
The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and $1.9\cdot10^{14}$ for ESRF-1 (see \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and especially in peak value. The number of photons per second after the multilayer monochromator is $2.9\cdot10^{15}$ for EBS and $1.9\cdot10^{15}$ for ESRF-1, considering that the energy bandwidth transmitted by a multilayer monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing, at this point, ideal reflectivity (100~\%) for the multilayers.

\begin{figure}\label{fig:FluxU18}
    \centering
    \includegraphics[width=0.95\textwidth]{GRAPHICS/fluxU18.png}
    \caption{Flux of the U18.3 undulator integrated over a square slit placed at 27.1~m from the ID centre with aperture of 1 and 4 mm.}
\end{figure}

To estimate the geometrical transmission of the beamline we have to consider which optical elements crop the beam and remove photons. This can be analyzed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h$=50~$\mu$m in horizontal and fully opened in vertical. For ESRF-1 its transmission is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their useful optical length onto a direction perpendicular to the optical axis. Using a mirror length of $L_v$=60~mm and $L_h$=26~mm for VFM and HFM, respectively, and incident angle of 15~mrad, the KB behaves as an aperture of  390 $\times$ 900 (H $\times$ V) $\mu$m$^2$. 
We get N.A. values of 2.7 $\times$ 4.9~$\mu$rad$^2$ (aperture size divided by the distance from the mirror to the VSS in H and to the source in V). The transmission coefficient is the ratio of the accepted N.A. over the beam divergence, resulting in 
4.1\% (H) and 40\% (V) for ESRF-1 and 7.1\% and 40\% for EBS. The overall transmission considers the product of H and V ratios, therefore 
0.21\% for ESRF-1 and 2.85\% for EBS, thus giving 
$4\cdot10^{12}$ photons/s for ESRF-1 and $8.3\cdot10^{13}$ photons/s for EBS. 
%\todo{check with references $10^9$ at 33.6 keV. I get: 1.6e+15 * 0.00295 * 0.7**3 = 1.62e+12. How closed are the slits? We used the undulator peak value over the whole ML bandwidth, that overestimates the integrated flux.}
%\todo{need to consider ML reflectivity for the 3 reflections, also undulator is not perfect, the horizontal slit may be closed to less than $50 \mu m$, also KB mirrors may not be fully illuminated and slope error losses due to initial ML mono are neglected}

One can also make some simple inferences about the beam coherence. The coherent fraction (CF) \cite{arxivCF} of the undulator beam is the occupation of the lowest coherent mode and will be discussed further in Section~\ref{level2}. An approximated estimation of CF can be obtained using Eqs.~\ref{eq:photon small sigmas} and \ref{eq:photon big sigmas}. The coherent fraction is the ratio of the phase space of the fully coherent (zero emittance) beam relative to the true beam. The CF in the horizontal and vertical planes are
\begin{equation}\label{eq:coherent fraction}
 CF_{(h,v)} = \frac{\sigma_u \sigma'_u}{\Sigma_{(h,v)} \Sigma'_{(h,v)}},
\end{equation}
and the total coherent fraction is the product of both (CF=CF$_h \times $CF$_v$). An estimation of the CF calculated in this way gives CF=0.12\% for the ESRF-1 and 2.73\% for EBS at 17 keV photon energy. The CF roughly indicates the ratio of the ``coherent photons'' over the total number of photons. It is a good indicator of the quality of a coherent beam (a fully coherent beam has CF=1 and an incoherent beam CF=0). The beamline optics act as a ``coherence filter'' by removing the ``bad'' or ``incoherent'' photons and in consequence increasing the CF. A complete cleaning is impossible because of the spatial overlapping of the different coherence modes. However, apertures centered at the optical axis would clean the beam from these undesired photons. If we compare the values of the beamline transmission due to the geometrical aperture  (0.21\% for ESRF-1 and 2.85\% for EBS) with the CF values, we remark that they are very close. This suggests that the beam at the focal position will be highly coherent because the fraction of the photons removed from the beam matches the CF value that quantifies the fraction of ``good'' or ``coherent'' photons.

To assess the focused beam size given by a coherent beam, one can consider the beamline in two sections: First, the ML monochromator and VSS that produce a coherent plane wave as the source (vertical) and VSS (horizontal) are both very far from the entrance aperture of the focusing optics. Second, the KB-system playing a dual role, acting as a focusing element of focal distance $f=5$~cm, but also cropping the beam by acting as a square aperture of roughly D$\times$D=400$\times$400~$\mu$m$^2$ (the length of the mirror projected in a plane perpendicular to the beam propagation).
%\todo{REMOVE: we obtained 900 in H, that will be reduced by 0.5 as propagated from 10 cm to 5 cm upstream of the sample, and 390 in V approximated by 400}.
This aperture has two effects: i) as discussed before, it removes a lot of off-axis photons increasing the CF to a value close to one, and ii) broadens of the focal spot because of coherent diffraction. The diffraction by a square aperture of a collapsing spherical wavefront has an intensity distribution proportional to $\sinc^2(k D x/2 f) \sinc^2(k D y / 2 f)$,
%\todo{include citation?}
where $k=2\pi/\lambda$, and $x$ ($y$) is the horizontal (vertical) coordinate. Considering that the FWHM of $\sinc(x)$ is approximately 2.78, one obtains a FWHM of 7.75$\times$7.75~nm$^2$ for the focal spot. This spot size is of the order or larger than that calculated geometrical considerations alone and therefore contributes significantly to the overall size of the focused beam. One can conclude that for the conditions in use this is a diffraction-limited beamline.

@@

%-------------------------------------------------------------------------
% 
%-------------------------------------------------------------------------
\subsection{Ray-tracing calculations}
\label{level1}
In this section, we perform ray-tracing calculations to study the effect of real optics (including aberrations and slope errors) upon the beam size and beamline transmission. The first part deals with the effect of aberrations and transmission due to geometrical considerations. The second part shows how the effect of slope errors of the KB mirrors degrade the focal spot.

Ray-tracing is a simulation method based on tracing some light rays along the optical system from the source to the image and retrieving the statistics of rays at the image to measure beam characteristics. For simple systems one can perform ray-tracing even by hand, selecting some principal rays that will define the location and envelope of the image. Using computers one can trace thousands of rays and calculate the position and divergence distribution of the rays using statistics. Rays are usually generated by Monte Carlo sampling the source characteristics. Ray-tracing is based on geometrical optics: a ray is a solution of the Helmholtz equation and travels in vacuum along a straight line. Rays are intercepted by the optical elements, which change their trajectory. For reflectors, this change of direction is given by the specular reflection laws. For refractors, the change of direction is defined by  Snell's law. Because Snell's law uses the refractive index which is wavelength-dependent, it is then interesting to assign additional attributes to each ray, such as the wavelength. This will permit us to calculate the refractive index that should be used for each ray. The characteristics of each ray can also be extended to include the electric field amplitude. This allows extending the pure geometrical tracing to include physical models that take into account the optical element reflectivity or transmission. This combination of using a geometrical model plus a physical model allows to simulate every element used in a synchrotron beamline, such as mirrors and combinations of them (e.g., KBs); lenses, compound refractive lenses, and transfocators; gratings of any type and shape, crystal systems, etc. Ray-tracing is a simple and extremely powerful technique for calculating the main characteristics of the photon beam (size, divergence, photon distribution) at every point of the beamline. 
%Although very powerful commercial ray-tracing packages are available in the market, they are generally ill-adapted for synchrotron simulations because they do not cover the particularities of X-ray devices and sources. For that purposes, 
Several packages created in the synchrotron community and are available, like SHADOW \cite{codeSHADOW}, RAY \cite{codeRAY}, McXtrace \cite{codeMCXTRACE} or XRT \cite{codeXRT}. For this study, we use SHADOW and its interface ShadowOUI \cite{codeSHADOWOUI}, available as a module of the OASYS suite.

@@

\subsection{Wave optics simulations}
\label{level2}

@@

% Table \ref{tab:SRW_results} compiles the beam sizes at focal position for mentioned simulations.

@@

%{\todo{more detailed comments on comparisons with filament beam results? For comparison also information on calculation time, number of electrons, ...}

% 
% Synchrotron radiation on the hard X-rays regime is hardly full coherent even for the so-called diffraction limited sources (see Table \ref{tab:Coherence}). To better describe the nature of the emitted radiation, one makes use of the partial coherent simulations using the \textsc{srw}, i.e. incoherent addition of the intensity from the emission from electrons sampled from the source's 6D phase-space \cite{Chubar2011b}. This is a natural extension from the method presented on the previous section.

>>>>

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF \cite{ID16A} in two scenarios. This beamline was built during the Upgrade Phase~I at ESRF (2009-2015). It uses one or several undulators as the source. During the period 2011-2018, it operated exploiting the ESRF-1 storage ring double-bend achromat (DBA) lattice with horizontal emittance about 4000~pm. In the second phase of the ESRF Upgrade, the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by \inred{one or two orders of magnitude}, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the X-ray sources will be much more coherent, especially in the horizontal direction.

ID16A is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. The beamline can operate at two fixed photon energies \inred{(at ca. 17 and 33 keV) }. The straight section in the magnet lattice houses four undulators, two 18~mm and another two of 22.4~mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

\begin{table}\label{tab:Undulators}
\centering
\caption{Parameters of the undulators available at the ID16A beamline: period ($\lambda_u$), length ($L$), number of identical undulators ($N$) and deflection parameter ($K$) values for \inred{two main photon energies}. The valuse are shown for the two ESRF magnetic lattices: the old ESRF-1, with $E_e = 6.04$~GeV), and the new EBS, with $E_e = 6.00$~GeV. $n$ is the emission harmonic in use for each particular photon energy.}

\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
name  & $\lambda_u$ [mm] & $L$ [m] & N &  $K$ (17050~eV) & $K$ (33600~eV) \\ \\
%\input{tableUndulator.txt}
 ESRF-1 U18.3  & 18.30  & 1.40  & 2 & 0.470(n=1) & 1.175(n=3)\\ 
 ESRF-1 U22.4  & 22.40  & 1.40  & 2 & 1.855(n=3) & 0.873(n=3)\\ \\
 EBS U18.3       & 18.30  & 1.40  & 2 &  0.437(n=1) & 1.156(n=3)\\ 
 EBS U22.4       & 22.40  & 1.40  & 2 &  1.836(n=3) & 0.852(n=3)\\ 
\end{tabular}%
}
\end{table}

@@

\begin{figure}\label{fig:ID16A}
    \centering
    \includegraphics[width=0.9\textwidth,clip=true,trim=50 85 50 85]{GRAPHICS/ID16A.pdf}
    \caption{Schematic of the ID16A beamline showing the horizontal (top) and vertical (bottom) planes and the position of the main optical elements. In the sketch ML stands for the focusing multilayer monochromator; VSS for virtual source slit; and KB for the Kirkpatrick-Baez set of mirrors. The sketch is drawn \inred{not to} scale.}
\end{figure}

@@

\begin{table}\label{tab:eBeam}
    \centering
    \caption{ \inred{Electron beam sizes for the ESRF-1 high-$\beta$ straight sections and the EBS straight sections. Values taken at the symmetry point of the straight section, where the insertion devices are placed}. Values are  from \cite{orangebook}, and updated for EBS with the lattice labelled S28D. 
    %For EBS, two (very close) design values are used, from Ref.~\cite{ESRF2014} labelled EBS-S28A, and those coming from further lattice refinement, labelled EBS-S28D. 
    %\todo{Remove ESRF-AT and EBS-S28A} 
    }
        \resizebox{\textwidth}{!}{%
        \begin{tabular}{ccccc}
         Lattice & $\sigma_e$ horizontal {[}$\mu$m{]} &  $\sigma_e'$ horizontal {[}$\mu$rad{]} & $\sigma_e$ vertical {[}$\mu$m{]} & $\sigma_e'$ vertical {[}$\mu$rad{]} \\ \\
         %\input{tableElectron.txt}
         ESRF-1  & 387.80  & 3.50  & 10.30 & 1.20 \\ 
         %ESRF-AT  & 414.97  & 3.43  & 10.31 & 1.16 \\ 
         %EBS-S28A  & 27.20  & 3.40  & 5.20 & 1.40 \\ 
         EBS     & 30.18  & 3.64  & 4.37 & 1.37 \\ 
    \end{tabular}
    }
\end{table}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$~eV (Table \ref{tab:Undulators}). Calculations were done using the two different storage ring lattices mentioned before: the (already dismantled) ESRF-1 where the undulators are in a high-$\beta$ straight section (larger horizontal emittance) and the new EBS (lower horizontal emittance) straight sections. The parameters for the current and the new lattice are shown in Tables \ref{tab:TableSources} and \ref{tab:eBeam}.
This section is divided into three subsections, that present three main levels of simulations: simple calculations that can be made by hand, ray-tracing simulations and wave optics simulations: sections  \ref{level0}, \ref{level1} section \ref{level2} respectively.

 %-------------------------------------------------------------------------
\subsection{Analytical calculations}
\label{level0}

@@

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \frac{1}{2  \pi}\sqrt{2 \lambda L}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization).  \inred{Eqs.~\ref{eq:photon small sigmas} are approximations, valid at the resonance, and under the assumptions that the undulator emission has Gaussian distribution in both angular and spatial coordinates, which is not the case as discussed below and in Ref.~\cite{arxivCF}} . The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source has sizes $\Sigma$ and divergences $\Sigma'$ given by:
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

The beam sizes at different positions in the beamline are estimated in different ways depending upon whether the beam is focused or not. If the beam is focused, its dimensions, to a first \inred{geometrical} approximation \inred{also neglecting diffraction effects}, are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam sizes in a position out of the focus the beam divergence is used. In an ideal {\'{e}}tendue-conserving optical system, where the Smith-Helmholz invariant applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$. \inred{It is, however, limited by the numerical aperture accepted by the optical elements due to finite size.}

@@

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$~eV for both the \inred{ESRF-1 (high-$\beta$)} and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50~$\mu$m.
    %(21.3 if approximating by the Gaussian $\sigma$).
    }
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF-1 (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                                 & Undulator     & VSS                      & Sample               & Undulator & VVS   & Sample       \\
    Position [m]:                          & 0.00                & 40.00                   & 185.00                  & 0.00    & 40.00   & 185.00       \\
    FWHM$_h$ [$\mu$m]:      & 977.19            & 404.00 (50.0)    & 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    FWHM$_v$ [$\mu$m]:      & 9.60                 & 478.7                  & 0.0052                  & 10.00   & 483.6   & 0.0054       \\
    \newline
    FWHM'$_h$ [$\mu$rad]:   & 26.93            & 65.14                   & 7800                     & 15.54   & 37.60   & 7800       \\
    FWHM'$_v$ [$\mu$rad]:   & 11.97            & 11.97                   & 9000                     & 12.09   & 12.09   & 9000        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons at the sample position. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most advanced and well-maintained codes used in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated using SRW over an aperture of $1 \times 1$~mm$^2$ placed at $d=27.1$~m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use.
The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and $1.9\cdot10^{14}$ for ESRF-1 (see \inred{Fig.~} \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and especially in peak value. The number of photons per second after the multilayer monochromator is $2.9\cdot10^{15}$ for EBS and $1.9\cdot10^{15}$ for ESRF-1, considering that the energy bandwidth transmitted by a multilayer monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing, at this point, ideal reflectivity (100~\%) for the multilayers.

\begin{figure}\label{fig:FluxU18}
    \centering
    \includegraphics[width=0.95\textwidth]{GRAPHICS/fluxU18.png}
    \caption{Flux of the U18.3 undulator integrated over a square slit \inred{(1$\times$1~mm$^2$)} placed at 27.1~m from the ID centre.}
\end{figure}

To estimate the geometrical transmission of the beamline we have to consider which optical elements crop the beam and remove photons. This can be analyzed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a \inred{slit (}VSS\inred{)} closed to $S_h$=50~$\mu$m in horizontal and fully opened in vertical. For ESRF-1 its transmission is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their useful optical length onto a direction perpendicular to the optical axis. Using a mirror length of $L_v$=60~mm and $L_h$=26~mm for VFM and HFM, respectively, and incident angle of 15~mrad, the KB behaves as an aperture of  390 $\times$ 900 (H $\times$ V) $\mu$m$^2$. 
We get N.A. values of 2.7 $\times$ 4.9~$\mu$rad$^2$ (aperture size divided by the distance from the mirror to the VSS in H and to the source in V). The transmission coefficient is the ratio of the accepted N.A. over the beam divergence, resulting in 
4.1\% (H) and 40\% (V) for ESRF-1 and 7.1\% and 40\% for EBS. The overall transmission considers the product of H and V ratios, therefore 
0.21\% for ESRF-1 and 2.85\% for EBS, thus giving 
$4\cdot10^{12}$ photons/s for ESRF-1 and $8.3\cdot10^{13}$ photons/s for EBS.

One can also make some simple inferences about the beam coherence. The coherent fraction (CF) of the undulator beam is the occupation of the lowest coherent mode \cite{arxivCF} and will be discussed further in Section~\ref{level2}. An approximated estimation of CF can be obtained using Eqs.~\ref{eq:photon small sigmas} and \ref{eq:photon big sigmas}. The coherent fraction is the ratio of the phase space of the fully coherent (zero emittance) beam relative to the true beam. The CF in the horizontal and vertical planes are
\begin{equation}\label{eq:coherent fraction}
 CF_{(h,v)} = \frac{\sigma_u \sigma'_u}{\Sigma_{(h,v)} \Sigma'_{(h,v)}},
\end{equation}
and the total coherent fraction is the product of both (CF=CF$_h \times $CF$_v$). An estimation of the CF calculated in this way gives CF=0.12\% for the ESRF-1 and 2.73\% for EBS at 17 keV photon energy. The CF roughly indicates the ratio of the ``coherent photons'' over the total number of photons. It is a good indicator of the quality of a coherent beam (a fully coherent beam has CF=1 and an incoherent beam CF=0). The beamline optics act as a ``coherence filter'' by removing the ``incoherent'' photons and in consequence increasing the CF. A complete cleaning is impossible because of the spatial overlapping of the different coherence modes. However, apertures centered at the optical axis would clean the beam from these undesired photons. If we compare the values of the beamline transmission due to the geometrical aperture  (0.21\% for ESRF-1 and 2.85\% for EBS) with the CF values, we remark that they are very close. This suggests that the beam at the focal position will be highly coherent because the fraction of the photons removed from the beam matches the CF value that quantifies the fraction of  ``coherent'' photons.

To assess the focused beam size given by a coherent beam, one can consider the beamline in two sections: First, the ML monochromator and VSS that produce a coherent plane wave as the source (vertical) and VSS (horizontal) are both very far from the entrance aperture of the focusing optics. Second, the KB-system playing a dual role, acting as a focusing element of focal distance $f=5$~cm, but also cropping the beam by acting as a square aperture of roughly D$\times$D=400$\times$400~$\mu$m$^2$ (the length of the mirror projected in a plane perpendicular to the beam propagation
\footnote{\inred{we obtained 900 $\mu$m in horizontal that is reduced by a factor 0.5 as propagated from 10 cm to 5 cm upstream of the sample, and 390 $\mu$m in vertical. These numbers are approximated by 400$\times$400$\mu$m$^2$ }}
).
This aperture has two effects: i) as discussed before, it removes a lot of off-axis photons increasing the CF to a value close to one, and ii) broadens of the focal spot because of coherent diffraction. The diffraction by a square aperture of a collapsing spherical wavefront has an intensity distribution proportional to $\sinc^2(k D x/2 f) \sinc^2(k D y / 2 f)$,
where $k=2\pi/\lambda$, and $x$ ($y$) is the horizontal (vertical) coordinate. Considering that the FWHM of $\sinc(x)$ is approximately 2.78, one obtains a FWHM of 7.75$\times$7.75~nm$^2$ for the focal spot. This spot size is of the order or larger than that calculated geometrical considerations alone and therefore contributes significantly to the overall size of the focused beam. One can conclude that for the conditions in use this is a diffraction-limited beamline.

@@

%-------------------------------------------------------------------------
\subsection{Ray-tracing calculations}
\label{level1}
In this section, we perform ray-tracing calculations to study the effect of real optics (including aberrations and slope errors) upon the beam size and beamline transmission. The first part deals with the effect of aberrations and transmission due to geometrical considerations. The second part shows how the effect of slope errors of the KB mirrors degrade the focal spot.

Ray-tracing is a simulation method based on tracing some light rays along the optical system from the source to the image and retrieving the statistics of rays at the image to measure beam characteristics. For simple systems one can perform ray-tracing even by hand, selecting some principal rays that will define the location and envelope of the image. Using computers one can trace thousands of rays and calculate the position and divergence distribution of the rays using statistics. Rays are usually generated by Monte Carlo sampling the source characteristics. Ray-tracing is based on geometrical optics: a ray is a solution of the Helmholtz equation and travels in vacuum along a straight line. Rays are intercepted by the optical elements, which change their trajectory. For reflectors, this change of direction is given by the specular reflection laws. For refractors, the change of direction is defined by  Snell's law. Because Snell's law uses the refractive index which is wavelength-dependent, it is then interesting to assign additional attributes to each ray, such as the wavelength. This will permit us to calculate the refractive index that should be used for each ray. The characteristics of each ray can also be extended to include the electric field amplitude\inred{, thus complementing} the pure geometrical tracing to include physical models that take into account the optical element reflectivity or transmission. This combination of using a geometrical model plus a physical model \inred{permits to simulate} every element used in a synchrotron beamline, such as mirrors and combinations of them (e.g., KBs); lenses, compound refractive lenses, and transfocators; gratings of any type and shape, crystal systems, etc. Ray-tracing is a simple and extremely powerful technique for calculating the main characteristics of the photon beam (size, divergence, photon distribution) at every point of the beamline. 
Several packages created in the synchrotron community and are available, like SHADOW \cite{codeSHADOW}, RAY \cite{codeRAY}, McXtrace \cite{codeMCXTRACE} or XRT \cite{codeXRT}. For this study, we use SHADOW and its interface ShadowOUI \cite{codeSHADOWOUI}, available as a module of the OASYS suite.

@@

 %-------------------------------------------------------------------------
\subsection{Wave optics simulations}
\label{level2}

@@

@@

\subsubsection{Full Wave optics simulation: partial coherence case by means of coherent mode decomposition}
\label{comsyl}
